\bf Appendix: Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-04-20 appendix.qmd
Updated Figure 1

```{r}
library(ggplot2)
library(dplyr)
library(quollr)
library(readr)
library(ggforce)
library(patchwork)
library(kableExtra)
library(tidyr)

theme_set(theme_linedraw() +
   theme(
     #aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "none", 
     legend.title = element_text(size=5), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm')
   )
)

interior_annotation <- function(label, position = c(0.92, 0.92), cex = 1, col="grey70") {
  annotation_custom(grid::textGrob(label = label,
      x = unit(position[1], "npc"), y = unit(position[2], "npc"),
      gp = grid::gpar(cex = cex, col=col)))
}
```

```{r}
#| label: scale-highd-data
# Center the data by subtracting the mean of each column
center_data <- function(data) {
  apply(data, 2, function(col) col - mean(col))
}

# Function to scale data manually
scale_data_manual <- function(data, type_col) {
  # Step 1: Center the data (mean 0)
  data_centered <- center_data(data |> select(-all_of(type_col)))

  # Step 2: Calculate the standard deviation of each dimension
  sds <- apply(data_centered, 2, sd)

  # Step 3: Scale each dimension to have the range [0, 1]
  data_scaled <- apply(data_centered, 2, function(col) col / max(abs(col)))

  # Step 4: Scale dimensions according to their variation
  # The dimension with the highest standard deviation is scaled to [-1, 1]
  # Other dimensions are scaled to smaller ranges based on their standard deviations
  max_sd <- max(sds)

  # Normalize the standard deviations to get scaling factors
  scaling_factors <- sds / max_sd

  for (i in seq_along(scaling_factors)) {
    data_scaled[, i] <- data_scaled[, i] * scaling_factors[i]
  }

  # Combine the scaled data with the 'type' column and return as a tibble
  data_scaled <- as_tibble(data_scaled) %>% 
    mutate(!!type_col := data[[type_col]])

  return(data_scaled)
}
```

```{r}
#| label: gen-axes

gen_axes <- function(proj, limits = 1, axis_pos_x = NULL, axis_pos_y = NULL, axis_labels, threshold) {

  axis_scale <- limits/6

  if (is.null(axis_pos_x)) {

    axis_pos_x <- -2/3 * limits

  }

  if (is.null(axis_pos_y)) {

    axis_pos_y <- -2/3 * limits

  }

  adj <- function(x, axis_pos) axis_pos + x * axis_scale
  axes <- data.frame(x1 = adj(0, axis_pos_x), 
                     y1 = adj(0, axis_pos_y), 
                     x2 = adj(proj[, 1], axis_pos_x), 
                     y2 = adj(proj[, 2], axis_pos_y))

  rownames(axes) <- axis_labels

  ## To remove axes
  axes <- axes |>
    mutate(distance = sqrt((x2 - x1)^2 + (y2 - y1)^2)) |>
    filter(distance >= threshold)

  theta <- seq(0, 2 * pi, length = 50)
  circle <- data.frame(c1 = adj(cos(theta), axis_pos_x), 
                       c2 = adj(sin(theta), axis_pos_y))

  return(list(axes = axes, circle = circle))

}
```

```{r}
#| label: solve-quad-fun

quad <- function(a = 3, b = 2 * a2, c = -(a2^2 + a1^2))
{
  a <- as.complex(a)
  answer <- c((-b + sqrt(b^2 - 4 * a * c)) / (2 * a),
              (-b - sqrt(b^2 - 4 * a * c)) / (2 * a))
  if(all(Im(answer) == 0)) answer <- Re(answer)
  if(answer[1] == answer[2]) return(answer[1])
  answer[answer>0] ## only positive 
}

@@

@@

```{r}
#| label: fig-hex-param
#| fig-cap: "The components of the hexagon grid illustrating notation."
#| out-height: 30%
#| fig-width: 2.5
#| fig-height: 3.3
#| fig-pos: H

@@

```{r}
#| label: read-pbmc-umap-layouts
# Read a variety of different NLDR representations of PBMC
# and plot them on same aspect ratio
umap_pbmc1 <- read_rds("data/pbmc3k/pbmc_scdeed_umap_n_neighbors_30_min_dist_0.3.rds")
umap_pbmc1 <- as_tibble(umap_pbmc1)
names(umap_pbmc1) <- c("UMAP1", "UMAP2")

@@

```{r}
#| label: combine-data-pbmc-umap

@@

```{r}
#| label: error-comp-pbmc-umap

@@

```{r}
#| label: fig-pbmc-mse-umap
#| fig-pos: H
#| out-height: 53%
#| fig-cap: "Assessing which of the 2 UMAP layouts with different hyper-parameter settings (n_neighbors: $30$, min_dist: $0.3$ (red); n_neighbors: $80$, min_dist: $0.5$ (blue)) on the PBMC data is the better representation using MSE for varying binwidth ($a_1$). Colour used for the lines and points in the top plot and in the scatterplots represents UMAP layout (a, b). Of the two, layout b is optimal across all binwidths making it the best choice."

@@

```{r}
#| label: published-pbmc-scdeed

@@

```{r}
#| label: hexbin-pbmc-scdeed
## Compute hexbin parameters
num_bins_x_pbmc <- 12
lim1 <- pbmc_scaled_obj_umap$lim1
lim2 <- pbmc_scaled_obj_umap$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

@@

```{r}
#| label: pbmc-umap-model-proj-scdeed

>>>>

```{r include=FALSE}
#| label: install-libraries
#| warning: false
#| echo: false

options(repos = c(CRAN = "https://cran.rstudio.com")) # Setup mirror

packages_to_check <- c("remotes", "tidyverse", "patchwork", "colorspace", "kableExtra", "conflicted", "ggforce")

for (pkg in packages_to_check) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message(paste("Installing package:", pkg))
    install.packages(pkg)
  } else {
    installed_version <- packageVersion(pkg)
    available_version <- tryCatch({
      utils::packageDescription(pkg)$Version
    }, error = function(e) NA) # Handle cases where package info isn't readily available

    if (!is.na(available_version) && installed_version < package_version(available_version)) {
      message(paste("A newer version of package", pkg, "is available. Updating..."))
      install.packages(pkg)
    } else {
      message(paste("Package", pkg, "is up-to-date (version", installed_version, ")."))
    }
  }
}

if (!requireNamespace("quollr", quietly = TRUE)) {
  remotes::install_github("JayaniLakshika/quollr")
}

@@

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(quollr)
library(tibble)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(patchwork)
library(colorspace)
library(kableExtra)
library(conflicted)
library(ggforce)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
```

```{r}
#| label: plot-theme
theme_set(theme_linedraw() +
   theme(
     #aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "bottom", 
     legend.title = element_blank(), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm')
   )
)
```

```{r}
#| label: import-scripts
source("scripts/additional_functions.R")
```

```{r}
#| label: code-setup
set.seed(20240110)
```

@@

```{r}
#| label: fig-hex-param
#| fig-cap: "The components of the hexagon grid illustrating notation."
#| out-width: 30%
#| fig-pos: H

@@

```{r}
#| label: read-pbmc-umap-layouts
#| eval: false
# Read a variety of different NLDR representations of PBMC
# and plot them on same aspect ratio
umap_pbmc1 <- read_rds("data/pbmc3k/pbmc_scdeed_umap_n_neighbors_30_min_dist_0.3.rds")
umap_pbmc1 <- as_tibble(umap_pbmc1)
names(umap_pbmc1) <- c("UMAP1", "UMAP2")

@@

```{r}
#| label: combine-data-pbmc-umap
#| eval: false

@@

```{r}
#| label: error-comp-pbmc-umap
#| eval: false

@@

```{r}
#| label: fig-pbmc-mse-umap
#| eval: false
#| fig-pos: H
#| out-height: 53%
#| fig-cap: "Assessing which of the 2 UMAP layouts with different hyper-parameter settings (n_neighbors: $30$, min_dist: $0.3$ (red); n_neighbors: $80$, min_dist: $0.5$ (blue)) on the PBMC data is the better representation using MSE for varying binwidth ($a_1$). Colour used for the lines and points in the top plot and in the scatterplots represents UMAP layout (a, b). Of the two, layout b is optimal across all binwidths making it the best choice."

@@

```{r}
#| label: published-pbmc-scdeed
#| eval: false

@@

```{r}
#| label: hexbin-pbmc-scdeed
#| eval: false
## Compute hexbin parameters
num_bins_x_pbmc <- 12
lim1 <- pbmc_scaled_obj_umap$lim1
lim2 <- pbmc_scaled_obj_umap$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

@@

```{r}
#| label: pbmc-umap-model-proj-scdeed
#| eval: false

### 2025-04-20 appendix.tex
Updated Figure 1

\includegraphics[width=1\textwidth,height=0.3\textheight]{appendix_files/figure-pdf/fig-hex-param-1.pdf}

@@

\includegraphics[width=1\textwidth,height=\textheight]{appendix_files/figure-pdf/fig-assign-data-1.pdf}

@@

\includegraphics[width=1\textwidth,height=0.3\textheight]{appendix_files/figure-pdf/fig-tri-param-1.pdf}

@@

Figure~\ref{fig-pbmc-mse-umap} shows layout from the original published
article along with that suggested by scDEED to be more accurate. The log
MSE vs binwidth plot illustrates that our approach would suggest that
scDEED is correct here, that layout b is more accurately reflecting the
cluster structure in the data.

\includegraphics[width=1\textwidth,height=0.53\textheight]{appendix_files/figure-pdf/fig-pbmc-mse-umap-1.pdf}

\caption{\label{fig-pbmc-mse-umap}Assessing which of the 2 UMAP layouts
with different hyper-parameter settings (n\_neighbors: \(30\),
min\_dist: \(0.3\) (red); n\_neighbors: \(80\), min\_dist: \(0.5\)
(blue)) on the PBMC data is the better representation using MSE for
varying binwidth (\(a_1\)). Colour used for the lines and points in the
top plot and in the scatterplots represents UMAP layout (a, b). Of the
two, layout b is optimal across all binwidths making it the best
choice.}

>>>>

\includegraphics[width=0.3\linewidth,height=\textheight,keepaspectratio]{appendix_files/figure-pdf/fig-hex-param-1.pdf}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{appendix_files/figure-pdf/fig-assign-data-1.pdf}

@@

\includegraphics[width=1\linewidth,height=0.3\textheight]{appendix_files/figure-pdf/fig-tri-param-1.pdf}

@@

\textbf{?@fig-pbmc-mse-umap} shows layout from the original published
article along with that suggested by scDEED to be more accurate. The log
MSE vs binwidth plot illustrates that our approach would suggest that
scDEED is correct here, that layout b is more accurately reflecting the
cluster structure in the data.
